---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit the latter. -->

## capemlGIS: tools to aid the generation of EML metadata for spatial vectors and rasters

### overview
  
This package extends the [CAPLTER/capeml](https://github.com/CAPLTER/capeml)
package to facilitate the creation of EML spatialRaster and spatialVector
metadata.


### installation

Install from GitHub (after installing the
[devtools](https://cran.r-project.org/web/packages/devtools/index.html) package:

```{r load capeml package, eval=FALSE}

devtools::install_github("CAPLTER/capemlGIS")
```

### getting started

Creating a EML dataset starts with the
[CAPLTER/capeml](https://github.com/CAPLTER/capeml) package, `capemlGIS` is
designed only to facilitate creating EML entities of type `spatialRaster` and
`spatialVector`. Users should please start with the `capeml` workflow,
including creating a `config.yaml` file that will feature project details that
`spatialRaster` and `spatialVector` require for processing.


### options

#### EML

This package defaults to the current version of EML. If desired, users can
switch to the previous version with `emld::eml_version("eml-2.1.1")`.

#### project naming

Most EML-generating functions in the capemlGIS package will create both
physical objects and EML references to those objects with the format:
`project-id`\_`object-name`\_`object-hash`\.`file-extension` (e.g.,
*664_site_map_5fb7b8d53d48010eab1a2e73db7f1941.kml*). The target object (e.g.,
site_map.kml) is renamed with the additional metadata and this object name is
referenced in the EML metadata. The exception to this approach are
spatialVectors where the hash of the file/object is not included in the new
object name. Note that the project-id is not passed to any of the functions,
and must exist in `config.yaml` (as `projectid`).

Project-naming functionality can be turned off by setting the `projectNaming`
option in `create_spatialRaster()` to FALSE. When set to FALSE, the object name
is not changed, and the file name of the object is included in the EML.

### tools to generate entity metadata

* `write_attributes()` creates a template as a csv file for supplying attribute
metadata for a spatial vector object that resides in the R environment
* `write_raster_factors()` creates a template as a csv file for supplying code
definition metadata for spatial rasters if raster values are categorical

### tools to create EML entities

* `create_spatialRaster()` creates a EML entity of type spatialRaster - see
  [vignette](https://caplter.github.io/capeml/articles/create_spatialRaster.html)
  for more detail
* `create_spatialVector()` creates a EML entity of type `spatialVector`. Output includes:
  + EML entity of type `spatialVector` that can be added to a EML dataset
  + input data written to a kml file with project naming (if selected).  Note
    that regardless of source type, all spatial vector input data are converted
    to type kml.

### overview: create a spatialVector

**example: myvector**

```{r myvector, eval=FALSE}

# Load spatial vector object, here reading an existing kml file but this
# could be by constructing a spatial object in R using any means.
msp_arthropod_locations <- sf::st_read("msp_arthropod_locations.kml")

# The 'Name' attribute of a kml file describes the points so be sure that this
# field is populated with meaninful data.
msp_arthropod_locations <- msp_arthropod_locations %>%
  mutate(Name = sampling_locations)

# Write the attributes to a file for editing. In the case of KML files, there
# are many fields that max exist but are not used (e.g., elevation); metadata
# for unused fields is not required and should be removed from the attributes
# metadata (taking advantage of the less stringent error checking by PASTA+ of
# spatial entities)
write_attributes(msp_arthropod_locations)

# Create a description of the data entity
msp_arthropod_locations_desc <- "Geospatial file (KML) detailing the
locations of ground-dwelling arthropod sampling sites in the McDowell
Sonoran Preserve, Scottsdale, Arizona. Polygons reflect the bounding box
(minimum and maximum extents) of paired (boundary-interior) sampling
locations. Precise sampling locations are available upon request to the CAP
LTER Data Manager."

# Generate EML object of type spatial vector, and write to file in KML format
msp_arthropod_locations_SV <- create_spatialVector(
  svname = msp_arthropod_locations,
  description = msp_arthropod_locations_desc
)

# The resulting spatialVector entity can be added to a EML dataset
```

