---
output: github_document
---

```{r libraries, echo=FALSE, message=FALSE, warning=FALSE}

library(kableExtra)
```

<!-- README.md is generated from README.Rmd. Please edit the latter. -->

## capemlGIS: tools to aid the generation of EML metadata for spatial vectors and rasters

### overview
  
This package extends the [CAPLTER/capeml](https://github.com/CAPLTER/capeml)
package to facilitate the creation of EML spatialRaster and spatialVector
metadata.

### installation

Install from GitHub (after installing the
[devtools](https://cran.r-project.org/web/packages/devtools/index.html) package:

```{r load capeml package, eval=FALSE}

devtools::install_github("CAPLTER/capemlGIS")
```

### options

#### EML

This package defaults to the current version of EML. If desired, users can
switch to the previous version with `emld::eml_version("eml-2.1.1")`.

#### project naming

Most EML-generating functions in the capemlGIS package will create both physical
objects and EML references to those objects with the format:
`project-id`\_`object-name`\_`object-hash`\.`file-extension` (e.g.,
*664_site_map_5fb7b8d53d48010eab1a2e73db7f1941.kml*). The target object (e.g.,
site_map.kml) is renamed with the additional metadata and this object name is
referenced in the EML metadata. The exception to this approach are
spatialVectors where the hash of the file/object is not included in the new
object name. Note that the project-id is not passed to any of the functions, and
must exist in the working R environment (as `projectid`).

Project-naming functionality can be turned off by setting the `projectNaming`
option in `create_spatialRaster()` to FALSE. When set to FALSE, the object name
is not changed, and the file name of the object is included in the EML.

### tools to generate entity metadata

* `write_attributes()` creates a template as a csv file for supplying attribute
metadata for a spatial vector object that resides in the R environment
* `write_raster_factors()` creates a template as a csv file for supplying code
definition metadata for spatial rasters if raster values are categorical

### tools to create EML entities

* `create_spatialRaster()` creates a EML entity of type spatialRaster - see
[vignette](https://caplter.github.io/capeml/articles/create_spatialRaster.html)
for more detail
* `create_spatialVector()` creates a EML entity of type spatialRaster
